ANOVA Comparison Simulations: Gaussian Data

### 2016-12-08 simulations/simulation.Rmd
Rerun sims with 1000 reps to get cleaner plots and power estimates

for individuals $i = 1, \dots, n_j$, $j = 1, \dots, J$.
$\beta_0$ is the coefficient for the standard normally-distributed baseline measurement $Y_{i0}$, $\beta_j$ is the mean effect of being at site $j$, $Z_{ij}$ is the treatment level, $\gamma_j$ is the effect of treatment at site $j$, and $\varepsilon_{ij}$ is an error term.
We will assume that $\beta_0 = 1$.
The observed $(Y_{ij0}, \varepsilon_{ij})$ are independent across $i$ and $j$.

Suppose there are three sites with $\beta_1 = 1, \beta_2 = 1.5,$ and $\beta_3 = 2$.
Assume that there are 16 individuals per site and treatment assignment is balanced, i.e. 8 people receive each treatment at each site.

@@

* Design 1: $\gamma_1 = \gamma_2 = \gamma_3 = 1$. This is the standard assumption of a constant, additive treatment effect.
* Design 2: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$. This is a constant, additive treatment effect at site 1, but no treatment effect at sites 2 and 3. This is a simplistic case of heterogeneous treatment effects.

@@

* ANOVA: we fit a linear model of response $Y_1$ on baseline $Y_0$, treatment $Z$, and a dummy for site.
* Stratified permutation: we permute treatment assignment within site, then take the difference in means between treated and control outcomes $Y_1$
* Differenced permutation: we do the same permutation procedure as the stratified permutation test, except we use the difference between outcome and baseline, $Y_1 - Y_0$
* Linear model (LM) permutation: we use the same stratified permutation procedure as above, except use the $t$-statistic for the coefficient on treatment in the linear regression of $Y_1$ on $Y_0$, $Z$, and site dummies
* Freedman-Lane test: see the other Rmd document for a full description of this procedure

Throughout our simulations, we first fix $Y_0$ and site ID.
Treatment $Z$ and the errors $\varepsilon$ are randomly drawn according to their respective distributions.
Then, $Y_1$ is constructed using the linear data-generating process above.
We regenerate $Z$, $\varepsilon$, and $Y_1$ 100 times for each design, then compute the empirical power of the five tests.

@@

```{r generate_sample}
generate_simulated_data <- function(gamma, effect, errors, n = c(16, 16, 16)){
  # Input:
  # gamma = the magnitude of the treatment effect
  # effect = "same effect" or "single site effect" - which sites have a tr effect > 0?
  # errors = "normal" or "heavy"
  # n = number of individuals at each site
  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), SITEID (stratum), site_effect (beta coefficient per individual), and epsilon (errors)

  SITEID <- rep(1:3, times = n)
  N <- sum(n)
  beta <- c(1, 1.5, 2)

@@

  # Generate covariates
  Y0 <- rnorm(N)
  Z <- rep(0:1, length.out = N)
  site_effect <- rep(beta, times = n)
  Y1 <- Y0 + gamma_vec*Z + site_effect + epsilon
  return(data.frame(Y1, Y0, Z, gamma_vec, SITEID, site_effect, epsilon))
}

generate_simulated_pvalues <- function(dataset, reps = 1e3){
  # Inputs:
  # dataset = a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), and SITEID (stratum)
  # Returns: a vector of p-values
  # first element is the p-value from the ANOVA
  # second element is the p-value from the stratified two-sample permutation test
  # third element is the p-value from the linear model test, permuting treatment
  # fourth element is the p-value from the Freedman-Lane linear model test, permuting residuals

  # ANOVA
  modelfit <- lm(Y1 ~ Y0 + Z + factor(SITEID), data = dataset)
  resanova <- summary(aov(modelfit))
  anova_pvalue <- resanova[[1]]["Z", "Pr(>F)"]

  # Stratified permutation test of Y1
  observed_diff_means <- mean(dataset$Y1[dataset$Z == 1]) - mean(dataset$Y1[dataset$Z == 0])
  diff_means_distr <- stratified_two_sample(group = dataset$Z, response = dataset$Y1, stratum = dataset$SITEID, reps = reps)
  perm_pvalue <- t2p(observed_diff_means, diff_means_distr, alternative = "two-sided")

  # Diffed permutation test of Y1-Y0
  dataset$diff <- dataset$Y1 - dataset$Y0
  observed_diff_means2 <- mean(dataset$diff[dataset$Z == 1]) - mean(dataset$diff[dataset$Z == 0])
  diff_means_distr2 <- stratified_two_sample(group = dataset$Z, response = dataset$diff, stratum = dataset$SITEID, reps = reps)
  perm_pvalue2 <- t2p(observed_diff_means2, diff_means_distr2, alternative = "two-sided")

  # Permutation of treatment in linear model
  observed_t1 <- summary(modelfit)[["coefficients"]]["Z", "t value"]
  lm1_t_distr <-  replicate(reps, {
    dataset$Z_perm <- permute_within_groups(dataset$Z, dataset$SITEID)
    lm1_perm <- lm(Y1 ~ Y0 + Z_perm + factor(SITEID), data = dataset)
    summary(lm1_perm)[["coefficients"]]["Z_perm", "t value"]
  })
  lm_pvalue <- t2p(observed_t1, lm1_t_distr, alternative = "two-sided")

  # Freedman-Lane linear model residual permutation
  lm2_no_tr <- lm(Y1~Y0 + factor(SITEID), data = dataset)
  lm2_resid <- residuals(lm2_no_tr)
  lm2_yhat <- fitted(lm2_no_tr)
  lm2_t_distr <-  replicate(reps, {
    lm2_resid_perm <- permute_within_groups(lm2_resid, dataset$SITEID)
    dataset$response_fl <-  lm2_yhat + lm2_resid_perm
    lm2_perm <- lm(response_fl ~ Y0 + Z + factor(SITEID), data = dataset)
    summary(lm2_perm)[["coefficients"]]["Z", "t value"]
  })
  fl_pvalue <- t2p(observed_t1, lm2_t_distr, alternative = "two-sided")

@@

Before testing for different kinds of effects, we begin checking that the tests have the correct level.
We follow the procedure described above, using an effect size of $\gamma = 0$ at all sites and using standard normal errors.
To have the correct level means that the test rate of rejection at level $\alpha$ is $\alpha 100\%$.
In other words, the p-values are uniformly distributed and the power curve should coincide with the line with slope $1$ through the origin.
Figure \ref{fig:design_null_plots} demonstrates that this is the case.
If anything, the differenced stratified permutation test has fewer than $\alpha 100\%$ false positives when using level $\alpha$.

@@

tmp <- generate_simulated_data(gamma = 0, effect = "same effect", errors = "normal")
design0_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design0_pvalues <- t(design0_pvalues)
colnames(design0_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design0_power <- apply(design0_pvalues, 2, compute_power)
```

@@

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
design1_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})

@@

## Design 2: Single site effect, normal errors

A similar pattern emerges: the simple stratified permutation test has low power, while the other four power curves roughly coincide.
The Freedman-Lane test may have the highest power for small p-values, but this could also just be noise.
All five power curves are closer to the line passing through the origin: since the effect is only present at one site, it is more difficult to detect.

@@

tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal")
design2_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design2_pvalues <- t(design2_pvalues)
colnames(design2_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design2_power <- apply(design2_pvalues, 2, compute_power)
```

```{r design2_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design2_power, "Effect of 1 at a Single Site, Normal Errors")

@@

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "heavy")
design3_pvalues <- replicate(100, {
  tmp$epsilon <- rt(nrow(tmp), df = 2)
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design3_pvalues <- t(design3_pvalues)
colnames(design3_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design3_power <- apply(design3_pvalues, 2, compute_power)
```

@@

## Design 4: single site effect, heavy-tailed errors

@@

tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "heavy")
design4_pvalues <- replicate(100, {
  tmp$epsilon <- rt(nrow(tmp), df = 2)
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design4_pvalues <- t(design4_pvalues)
colnames(design4_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design4_power <- apply(design4_pvalues, 2, compute_power)
```

```{r design4_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design4_power, "Effect of 1 at a Single Site, Heavy-tailed Errors")

@@

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
design5_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- 0.25*tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})

@@

Suppose that instead of having 16 individuals per site, we distribute them unequally across sites:
site 1 has 8 patients, site 2 has 16 patients, and site 3 has 24 patients.
This imbalance does not violate any assumptions of the ANOVA.
However, it may reduce power if the effect is concentrated in sites with fewer patients.

@@

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal", n = c(8, 16, 24))
design6_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})

@@

## Design 7: Single site effect, normal errors

There is very low power to detect the effect here, since in the simulated data, the site with only 8 patients was the only site with a nonzero effect of treatment.
The treatment effect is on the same order of magnitude as the error variance.

@@

tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal", n = c(8, 16, 24))
design7_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design7_pvalues <- t(design7_pvalues)
colnames(design7_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design7_power <- apply(design7_pvalues, 2, compute_power)
```

```{r design7_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design7_power, "Effect of 1 at a Single Site, Normal Errors")

@@

One assumption of ANOVA is that errors are homoskedastic: they have the same variance across groups.
Suppose that this does not hold in our data.
Imagine we have three sites with 16 patients each, but they have different error distributions:
site 1 has standard normal errors,
site 2 has normally distributed errors with mean 0 and variance 2,
and site 3 has errors with a $t$-distribution on 2 degrees of freedom.

@@

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal", n = rep(nn, 3))
design8_pvalues <- replicate(100, {
  tmp$epsilon[tmp$SITEID == 1] <- rnorm(nn)
  tmp$epsilon[tmp$SITEID == 2] <- rnorm(nn, sd = sqrt(2))
  tmp$epsilon[tmp$SITEID == 3] <- rt(nn, df = 2)
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})

@@

## Design 9: heteroskedastic errors, single site effect

@@

tmp <- generate_simulated_data(gamma = 1, effect = "single site effect", errors = "normal", n = rep(nn, 3))
design9_pvalues <- replicate(100, {
  tmp$epsilon[tmp$SITEID == 1] <- rnorm(nn)
  tmp$epsilon[tmp$SITEID == 2] <- rnorm(nn, sd = sqrt(2))
  tmp$epsilon[tmp$SITEID == 3] <- rt(nn, df = 2)
  tmp$Z <- permute_within_groups(tmp$Z, tmp$SITEID)
  tmp$Y1 <- tmp$Y0 + tmp$site_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design9_pvalues <- t(design9_pvalues)
colnames(design9_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design9_power <- apply(design9_pvalues, 2, compute_power)
```

```{r design9_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design9_power, "Effect of 1 at a Single Site, Heteroskedastic Errors")

@@

powers <- list(design1_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design1_power)), "Errors" = rep("Gaussian", nrow(design1_power))), 
               design2_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Site Effect", nrow(design2_power)), "Errors" = rep("Gaussian", nrow(design2_power))), 
               design3_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design3_power)), "Errors" = rep("Heavy-tailed", nrow(design3_power))), 
               design4_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Site Effect", nrow(design4_power)), "Errors" = rep("Heavy-tailed", nrow(design4_power))))

@@

>>>>

for individuals $i = 1, \dots, n_j$, $j = 1, \dots, J$.
$\beta_0$ is the coefficient for the standard normally-distributed baseline measurement $Y_{i0}$, $\beta_j$ is the mean effect of being in stratum $j$, $Z_{ij}$ is the treatment level, $\gamma_j$ is the effect of treatment in stratum $j$, and $\varepsilon_{ij}$ is an error term.
We will assume that $\beta_0 = 1$.
The observed $(Y_{ij0}, \varepsilon_{ij})$ are independent across $i$ and $j$.

Suppose there are three strata with $\beta_1 = 1, \beta_2 = 1.5,$ and $\beta_3 = 2$.
Assume that there are 16 individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment at each stratum.

@@

* Design 1: $\gamma_1 = \gamma_2 = \gamma_3 = 1$. This is the standard assumption of a constant, additive treatment effect.
* Design 2: $\gamma_1 = \gamma > 0$, $\gamma_2 = \gamma_3 = 0$. This is a constant, additive treatment effect in stratum 1, but no treatment effect in stratums 2 and 3. This is a simplistic case of heterogeneous treatment effects.

@@

* ANOVA: we fit a linear model of response $Y_1$ on baseline $Y_0$, treatment $Z$, and a dummy for stratum.
* Stratified permutation: we permute treatment assignment within stratum, then take the difference in means between treated and control outcomes $Y_1$
* Differenced permutation: we do the same permutation procedure as the stratified permutation test, except we use the difference between outcome and baseline, $Y_1 - Y_0$
* Linear model (LM) permutation: we use the same stratified permutation procedure as above, except use the $t$-statistic for the coefficient on treatment in the linear regression of $Y_1$ on $Y_0$, $Z$, and stratum dummies
* Freedman-Lane test: see the other Rmd document for a full description of this procedure

Throughout our simulations, we first fix $Y_0$ and stratum ID.
Treatment $Z$ and the errors $\varepsilon$ are randomly drawn according to their respective distributions.
Then, $Y_1$ is constructed using the linear data-generating process above.
We regenerate $Z$, $\varepsilon$, and $Y_1$ 100 times for each design, then compute the empirical power of the five tests.

@@

```{r generate_sample}
generate_simulated_data <- function(gamma, effect, errors, n = c(16, 16, 16)){
  # Input:
  # gamma = the magnitude of the treatment effect
  # effect = "same effect" or "single stratum effect" - which strata have a tr effect > 0?
  # errors = "normal" or "heavy"
  # n = number of individuals at each stratum
  # Returns: a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), gamma_vec (treatment effect per individual), stratumID (stratum), stratum_effect (beta coefficient per individual), and epsilon (errors)

  stratumID <- rep(1:3, times = n)
  N <- sum(n)
  beta <- c(1, 1.5, 2)

@@

  # Generate covariates
  Y0 <- rnorm(N)
  Z <- rep(0:1, length.out = N)
  stratum_effect <- rep(beta, times = n)
  Y1 <- Y0 + gamma_vec*Z + stratum_effect + epsilon
  return(data.frame(Y1, Y0, Z, gamma_vec, stratumID, stratum_effect, epsilon))
}

generate_simulated_pvalues <- function(dataset, reps = 1e3){
  # Inputs:
  # dataset = a dataframe containing columns named Y1 (response), Y0 (baseline), Z (treatment), and stratumID (stratum)
  # Returns: a vector of p-values
  # first element is the p-value from the ANOVA
  # second element is the p-value from the stratified two-sample permutation test
  # third element is the p-value from the linear model test, permuting treatment
  # fourth element is the p-value from the Freedman-Lane linear model test, permuting residuals

  # ANOVA
  modelfit <- lm(Y1 ~ Y0 + Z + factor(stratumID), data = dataset)
  resanova <- summary(aov(modelfit))
  anova_pvalue <- resanova[[1]]["Z", "Pr(>F)"]

  # Stratified permutation test of Y1
  observed_diff_means <- mean(dataset$Y1[dataset$Z == 1]) - mean(dataset$Y1[dataset$Z == 0])
  diff_means_distr <- stratified_two_sample(group = dataset$Z, response = dataset$Y1, stratum = dataset$stratumID, reps = reps)
  perm_pvalue <- t2p(observed_diff_means, diff_means_distr, alternative = "two-sided")

  # Diffed permutation test of Y1-Y0
  dataset$diff <- dataset$Y1 - dataset$Y0
  observed_diff_means2 <- mean(dataset$diff[dataset$Z == 1]) - mean(dataset$diff[dataset$Z == 0])
  diff_means_distr2 <- stratified_two_sample(group = dataset$Z, response = dataset$diff, stratum = dataset$stratumID, reps = reps)
  perm_pvalue2 <- t2p(observed_diff_means2, diff_means_distr2, alternative = "two-sided")

  # Permutation of treatment in linear model
  observed_t1 <- summary(modelfit)[["coefficients"]]["Z", "t value"]
  lm1_t_distr <-  replicate(reps, {
    dataset$Z_perm <- permute_within_groups(dataset$Z, dataset$stratumID)
    lm1_perm <- lm(Y1 ~ Y0 + Z_perm + factor(stratumID), data = dataset)
    summary(lm1_perm)[["coefficients"]]["Z_perm", "t value"]
  })
  lm_pvalue <- t2p(observed_t1, lm1_t_distr, alternative = "two-sided")

  # Freedman-Lane linear model residual permutation
  lm2_no_tr <- lm(Y1~Y0 + factor(stratumID), data = dataset)
  lm2_resid <- residuals(lm2_no_tr)
  lm2_yhat <- fitted(lm2_no_tr)
  lm2_t_distr <-  replicate(reps, {
    lm2_resid_perm <- permute_within_groups(lm2_resid, dataset$stratumID)
    dataset$response_fl <-  lm2_yhat + lm2_resid_perm
    lm2_perm <- lm(response_fl ~ Y0 + Z + factor(stratumID), data = dataset)
    summary(lm2_perm)[["coefficients"]]["Z", "t value"]
  })
  fl_pvalue <- t2p(observed_t1, lm2_t_distr, alternative = "two-sided")

@@

Before testing for different kinds of effects, we begin checking that the tests have the correct level.
We follow the procedure described above, using an effect size of $\gamma = 0$ at all strata and using standard normal errors.
To have the correct level means that the test rate of rejection at level $\alpha$ is $\alpha 100\%$.
In other words, the p-values are uniformly distributed and the power curve should coincide with the line with slope $1$ through the origin.
Figure \ref{fig:design_null_plots} demonstrates that this is the case.
If anything, the differenced stratified permutation test has fewer than $\alpha 100\%$ false positives when using level $\alpha$.

@@

tmp <- generate_simulated_data(gamma = 0, effect = "same effect", errors = "normal")
design0_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design0_pvalues <- t(design0_pvalues)
colnames(design0_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design0_power <- apply(design0_pvalues, 2, compute_power)
```

@@

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
design1_pvalues <- replicate(1000, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})

@@

## Design 2: Single stratum effect, normal errors

A similar pattern emerges: the simple stratified permutation test has low power, while the other four power curves roughly coincide.
The Freedman-Lane test may have the highest power for small p-values, but this could also just be noise.
All five power curves are closer to the line passing through the origin: since the effect is only present at one stratum, it is more difficult to detect.

@@

tmp <- generate_simulated_data(gamma = 1, effect = "single stratum effect", errors = "normal")
design2_pvalues <- replicate(1000, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design2_pvalues <- t(design2_pvalues)
colnames(design2_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design2_power <- apply(design2_pvalues, 2, compute_power)
```

```{r design2_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design2_power, "Effect of 1 at a Single stratum, Normal Errors")

@@

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "heavy")
design3_pvalues <- replicate(1000, {
  tmp$epsilon <- rt(nrow(tmp), df = 2)
  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design3_pvalues <- t(design3_pvalues)
colnames(design3_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design3_power <- apply(design3_pvalues, 2, compute_power)
```

@@

## Design 4: single stratum effect, heavy-tailed errors

@@

tmp <- generate_simulated_data(gamma = 1, effect = "single stratum effect", errors = "heavy")
design4_pvalues <- replicate(1000, {
  tmp$epsilon <- rt(nrow(tmp), df = 2)
  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design4_pvalues <- t(design4_pvalues)
colnames(design4_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design4_power <- apply(design4_pvalues, 2, compute_power)
```

```{r design4_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design4_power, "Effect of 1 at a Single stratum, Heavy-tailed Errors")

@@

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal")
design5_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
  tmp$Y1 <- 0.25*tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})

@@

Suppose that instead of having 16 individuals per stratum, we distribute them unequally across strata:
stratum 1 has 8 patients, stratum 2 has 16 patients, and stratum 3 has 24 patients.
This imbalance does not violate any assumptions of the ANOVA.
However, it may reduce power if the effect is concentrated in strata with fewer patients.

@@

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal", n = c(8, 16, 24))
design6_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})

@@

## Design 7: Single stratum effect, normal errors

There is very low power to detect the effect here, since in the simulated data, the stratum with only 8 patients was the only stratum with a nonzero effect of treatment.
The treatment effect is on the same order of magnitude as the error variance.

@@

tmp <- generate_simulated_data(gamma = 1, effect = "single stratum effect", errors = "normal", n = c(8, 16, 24))
design7_pvalues <- replicate(100, {
  tmp$epsilon <- rnorm(nrow(tmp))
  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design7_pvalues <- t(design7_pvalues)
colnames(design7_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design7_power <- apply(design7_pvalues, 2, compute_power)
```

```{r design7_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design7_power, "Effect of 1 at a Single stratum, Normal Errors")

@@

One assumption of ANOVA is that errors are homoskedastic: they have the same variance across groups.
Suppose that this does not hold in our data.
Imagine we have three strata with 16 patients each, but they have different error distributions:
stratum 1 has standard normal errors,
stratum 2 has normally distributed errors with mean 0 and variance 2,
and stratum 3 has errors with a $t$-distribution on 2 degrees of freedom.

@@

tmp <- generate_simulated_data(gamma = 1, effect = "same effect", errors = "normal", n = rep(nn, 3))
design8_pvalues <- replicate(100, {
  tmp$epsilon[tmp$stratumID == 1] <- rnorm(nn)
  tmp$epsilon[tmp$stratumID == 2] <- rnorm(nn, sd = sqrt(2))
  tmp$epsilon[tmp$stratumID == 3] <- rt(nn, df = 2)
  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})

@@

## Design 9: heteroskedastic errors, single stratum effect

@@

tmp <- generate_simulated_data(gamma = 1, effect = "single stratum effect", errors = "normal", n = rep(nn, 3))
design9_pvalues <- replicate(100, {
  tmp$epsilon[tmp$stratumID == 1] <- rnorm(nn)
  tmp$epsilon[tmp$stratumID == 2] <- rnorm(nn, sd = sqrt(2))
  tmp$epsilon[tmp$stratumID == 3] <- rt(nn, df = 2)
  tmp$Z <- permute_within_groups(tmp$Z, tmp$stratumID)
  tmp$Y1 <- tmp$Y0 + tmp$stratum_effect + tmp$gamma_vec*tmp$Z + tmp$epsilon
  generate_simulated_pvalues(tmp)
})
design9_pvalues <- t(design9_pvalues)
colnames(design9_pvalues) <- c("ANOVA", "Stratified Permutation", "Differenced Permutation", "LM Permutation", "Freedman-Lane")
design9_power <- apply(design9_pvalues, 2, compute_power)
```

```{r design9_plot, fig.height = 4, fig.align="center"}
plot_power_curves(design9_power, "Effect of 1 at a Single stratum, Heteroskedastic Errors")

@@

powers <- list(design1_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design1_power)), "Errors" = rep("Gaussian", nrow(design1_power))), 
               design2_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Stratum Effect", nrow(design2_power)), "Errors" = rep("Gaussian", nrow(design2_power))), 
               design3_power %>% as.data.frame() %>% mutate("Treatment" = rep("Constant Additive Effect", nrow(design3_power)), "Errors" = rep("Heavy-tailed", nrow(design3_power))), 
               design4_power %>% as.data.frame() %>% mutate("Treatment" = rep("Single Stratum Effect", nrow(design4_power)), "Errors" = rep("Heavy-tailed", nrow(design4_power))))

@@

```{r power_summary_tab, results = "asis"}
summary05 <- t(sapply(powers, function(x) x[5, c(6, 7, 1:5)]))
summarytab <- xtable(summary05, align = "rp{1.15in}|p{0.7in}|p{0.6in}p{0.8in}p{0.8in}p{0.8in}p{0.75in}",
       caption = "Empirical power at level $0.05$ for Gaussian simulated data",
       label = "tab:normal_power") 
print(summarytab, type="latex", file="../ms/fig/normal_simulation_power_summary.tex")
summarytab
```
